---
title: "Online Textbook"
author: "Ayisha - TRU- T00727585"
date: "2025-04-04"
output: html_document
---

```{r}
# Load required libraries
library(tidyverse)
library(janitor)



# Read the Online textbook survey
online_df <- read.csv("BIOL 1110 Survey on Open Online Textbooks.csv", header = TRUE, stringsAsFactors = FALSE)
  #clean_names()

# Read the Traditional textbook survey
traditional_df <- read.csv("BIOL 1110 Survey on Traditional Textbooks.csv",header = TRUE, stringsAsFactors = FALSE) 
  #clean_names()

```


```{r}
online_df <- online_df %>%
  mutate(textbook_type = "Online")

traditional_df <- traditional_df %>%
  mutate(textbook_type = "Traditional")

```

```{r}
names(online_df)
```
```{r}
names(traditional_df)
```

```{r}
dim(online_df)
```

```{r}
dim(traditional_df)
```



```{r}
# Bind the datasets together
combined_df <- bind_rows(online_df, traditional_df)

```

```{r}
print(combined_df)
```


```{r}
dim(combined_df)
```

```{r}
glimpse(combined_df)
table(combined_df$textbook_type)

```

```{r}
table(combined_df$Q5_Textbook_Usage_Frequency)
table(combined_df$Q1_Preferred_Study_Mode)
table(combined_df$Q2_Textbook_Type_Used)
table(combined_df$Q3_Prior_Open_Textbook_Experience)
table(combined_df$Q4_Clear_Explanations)
table(combined_df$Q4_Understanding_Difficult_Topics)
table(combined_df$Q4_Useful_Examples)
table(combined_df$Q4_Overall_Success_Contribution )
table(combined_df$Q5_Textbook_Usage_Frequency )
table(combined_df$Q6_Textbook_Effectiveness_Assignments_Studying)
table(combined_df$Q7_Engagement_Text_Readability)
table(combined_df$Q7_Engagement_Visual_Design )
table(combined_df$Q7_Engagement_Navigation_Ease )
table(combined_df$Q8_Textbook_Enhancement_Course_Connection)
table(combined_df$Q9_Textbook_Contribution_Retention  )
table(combined_df$Q10_Used_Interactive_Features)
table(combined_df$Q12_Interactive_Feature_Usefulness)
table(combined_df$Q13_Effectiveness_Understanding)
table(combined_df$Q13_Effectiveness_Retention)
table(combined_df$Q13_Effectiveness_ApplyingConcepts)
table(combined_df$Q13_Effectiveness_Reviewing)
table(combined_df$Q13_Effectiveness_Other)
table(combined_df$Q15_Textbook_Accessibility)
table(combined_df$Q16_Textbook_Preference_Future)
table(combined_df$Q17_Textbook_Distracting)
table(combined_df$Q18overall_textbook_satisfaction)
table(combined_df$textbook_type)




```

```{r}
summary(combined_df)
```


```{r}



# 🧭 Define actual factor levels based on your `table()` results

# Q5 + Q11 Frequency scale (includes "A few times a week")
likert_frequency_levels <- c("Never", "Rarely", "A few times a week", "Weekly", "Daily")

# Q4 Agreement
likert_agreement_levels <- c("Disagree", "Neutral", "Agree", "Strongly Agree")

# Q6 (Effectiveness w/ “Somewhat effective”)
likert_effectiveness_levels_simple <- c("Not at all effective", "Somewhat effective", "Very effective")

# Q7 (Engagement: full 5-point scale)
likert_engagement_levels <- c("Very Disengaging", "Disengaging", "Neutral", "Engaging", "Very Engaging")

# Q8–Q9 (Contribution: Includes “Very much”, “Moderately”, etc.)
likert_contribution_levels <- c("Not at all", "Slightly", "Moderately", "Very much", "Extremely")

# Q12 (Usefulness)
likert_usefulness_levels <- c("Not useful", "Slightly useful", "Moderately useful", "Very useful", "Extremely useful")

# Q13 (Effectiveness tasks)
likert_effectiveness_tasks <- c("Not Effective", "Slightly Effective", "Moderately Effective", "Very Effective", "Extremely Effective")

# Q15 Accessibility
likert_accessibility_levels <- c("Not Accessible", "Slightly Accessible", "Moderately Accessible", "Very Accessible", "Extremely Accessible")

# Q16 Preference (no recoding needed — categorical)
# Q17 Distraction
likert_distraction_levels <- c("Not at all distracting", "Slightly Distracting", "Moderately Distracting", "Very Distracting", "Extremely Distracting")

# Q18 Satisfaction
likert_satisfaction_levels <- c("Not Satisfied", "Slightly Satisfied", "Moderately Satisfied", "Very Satisfied", "Extremely Satisfied")

# ✅ Recode all applicable columns
combined_df <- combined_df %>%
  mutate(
    # Agreement block
    Q4_Clear_Explanations = factor(Q4_Clear_Explanations, levels = likert_agreement_levels, ordered = TRUE),
    Q4_Understanding_Difficult_Topics = factor(Q4_Understanding_Difficult_Topics, levels = likert_agreement_levels, ordered = TRUE),
    Q4_Useful_Examples = factor(Q4_Useful_Examples, levels = likert_agreement_levels, ordered = TRUE),
    Q4_Overall_Success_Contribution = factor(Q4_Overall_Success_Contribution, levels = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"), ordered = TRUE),

    # Usage frequency (Q5 and Q11)
    Q5_Textbook_Usage_Frequency = factor(Q5_Textbook_Usage_Frequency, levels = likert_frequency_levels, ordered = TRUE),
    Q11_Interactive_Feature_Usage_Frequency = factor(Q11_Interactive_Feature_Usage_Frequency, levels = likert_frequency_levels, ordered = TRUE),

    # Effectiveness (Q6 simplified scale)
    Q6_Textbook_Effectiveness_Assignments_Studying = factor(Q6_Textbook_Effectiveness_Assignments_Studying, levels = likert_effectiveness_levels_simple, ordered = TRUE),

    # Engagement (Q7)
    Q7_Engagement_Text_Readability = factor(Q7_Engagement_Text_Readability, levels = likert_engagement_levels, ordered = TRUE),
    Q7_Engagement_Visual_Design = factor(Q7_Engagement_Visual_Design, levels = likert_engagement_levels, ordered = TRUE),
    Q7_Engagement_Navigation_Ease = factor(Q7_Engagement_Navigation_Ease, levels = likert_engagement_levels, ordered = TRUE),

    # Enhancement/contribution (Q8, Q9)
    Q8_Textbook_Enhancement_Course_Connection = factor(Q8_Textbook_Enhancement_Course_Connection, levels = likert_contribution_levels, ordered = TRUE),
    Q9_Textbook_Contribution_Retention = factor(Q9_Textbook_Contribution_Retention, levels = likert_contribution_levels, ordered = TRUE),

    # Usefulness (Q12)
    Q12_Interactive_Feature_Usefulness = factor(Q12_Interactive_Feature_Usefulness, levels = likert_usefulness_levels, ordered = TRUE),

    # Effectiveness on tasks (Q13)
    Q13_Effectiveness_Understanding = factor(Q13_Effectiveness_Understanding, levels = likert_effectiveness_tasks, ordered = TRUE),
    Q13_Effectiveness_Retention = factor(Q13_Effectiveness_Retention, levels = likert_effectiveness_tasks, ordered = TRUE),
    Q13_Effectiveness_ApplyingConcepts = factor(Q13_Effectiveness_ApplyingConcepts, levels = likert_effectiveness_tasks, ordered = TRUE),
    Q13_Effectiveness_Reviewing = factor(Q13_Effectiveness_Reviewing, levels = likert_effectiveness_tasks, ordered = TRUE),
    Q13_Effectiveness_Other = factor(Q13_Effectiveness_Other, levels = likert_effectiveness_tasks, ordered = TRUE),

    # Accessibility (Q15)
    Q15_Textbook_Accessibility = factor(Q15_Textbook_Accessibility, levels = likert_accessibility_levels, ordered = TRUE),

    # Distraction (Q17)
    Q17_Textbook_Distracting = factor(Q17_Textbook_Distracting, levels = likert_distraction_levels, ordered = TRUE),

    # Satisfaction (Q18)
    Q18overall_textbook_satisfaction = factor(Q18overall_textbook_satisfaction, levels = likert_satisfaction_levels, ordered = TRUE)
  )

# ✅ Now all Likert variables are correctly ordered and ready for analysis!


```


```{r}

summary(combined_df)
```


```{r}

# For satisfaction comparison
combined_df_clean <- combined_df %>%
  filter(!is.na(Q18overall_textbook_satisfaction))

# Satisfaction by textbook type
table(combined_df$Q18overall_textbook_satisfaction, combined_df$textbook_type)

# Or for visual % comparison
prop.table(table(combined_df$Q18overall_textbook_satisfaction, combined_df$textbook_type), margin = 2)

```
```{r}
library(ggplot2)
png("satisfaction_plot.png", width = 1200, height = 800, res = 300)
ggplot(combined_df_clean, aes(x = textbook_type, fill = Q18overall_textbook_satisfaction)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()
dev.off()
```

```{r}
combined_df_clean <- combined_df %>%
  filter(!is.na(Q18overall_textbook_satisfaction))

wilcox.test(as.numeric(Q18overall_textbook_satisfaction) ~ textbook_type, data = combined_df_clean)

```
Since p > 0.05, we fail to reject the null hypothesis

This means:
There is no statistically significant difference in overall satisfaction between online and traditional textbook users (based on your data)


The comparison of overall textbook satisfaction between students who used online versus traditional textbooks revealed broadly similar patterns, with no statistically significant difference (p = 0.396). However, the proportional bar chart shows that a greater percentage of students using online textbooks reported being “Very Satisfied” compared to those using traditional materials.

While “Moderately Satisfied” was the most common response across both groups, traditional users tended to cluster more heavily in that category, whereas online users showed greater variation—including a higher percentage in both “Very Satisfied” and “Slightly Satisfied” categories.

This suggests that, although satisfaction levels may not differ significantly on average, online textbooks are perceived more positively by some students, and may offer unique benefits in terms of flexibility, accessibility, and engagement that resonate with certain learners.

These findings support the potential of open online textbooks as a viable and potentially preferred alternative, especially when thoughtfully integrated into course design.


***********************************************************************************************************

Step 2: Analyze Textbook Accessibility
If students felt online textbooks were more accessible, that supports your case strongly.

Why Accessibility?
It’s directly related to usability and inclusiveness

Accessibility is one of the core promises of open educational resources (OER)

It helps to move the argument beyond just "satisfaction"
```{r}
access_df <- combined_df %>%
  filter(!is.na(Q15_Textbook_Accessibility))

```

```{r}
ggplot(access_df, aes(x = textbook_type, fill = Q15_Textbook_Accessibility)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()

```

```{r}
# Save plot as PNG
png("accessibility_plot.png", width = 1200, height = 800, res = 300)

ggplot(access_df, aes(x = textbook_type, fill = Q15_Textbook_Accessibility)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()

dev.off()

```

```{r}
wilcox.test(as.numeric(Q15_Textbook_Accessibility) ~ textbook_type, data = access_df)

```

The analysis revealed a statistically significant difference in perceived textbook accessibility between students using online versus traditional textbooks (Wilcoxon rank-sum test, W = 768.5, p = 0.013). Students who used online textbooks reported significantly higher levels of accessibility, with a greater proportion selecting “Very Accessible” or “Extremely Accessible” compared to their peers using traditional print materials.

This suggests that online textbooks are not only a convenient alternative but may also be better suited to supporting flexible, inclusive, and user-friendly learning environments. The enhanced accessibility of digital formats supports the broader goal of equitable education and strengthens the case for adopting open online textbooks in post-secondary biology education.


*****************************************************************************************************

Step 3: Analyze Textbook Effectiveness for Assignments/Studying
Question: Q6_Textbook_Effectiveness_Assignments_Studying

This one evaluates how well students felt the textbook helped them with coursework, which is a practical angle in favor of online use.


```{r}
effectiveness_df <- combined_df %>%
  filter(!is.na(Q6_Textbook_Effectiveness_Assignments_Studying))

```

```{r}
# Save plot as PNG
png("effectiveness_plot.png", width = 1200, height = 800, res = 300)
ggplot(effectiveness_df, aes(x = textbook_type, fill = Q6_Textbook_Effectiveness_Assignments_Studying)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()
dev.off()
```

```{r}
library(stringr)

# Wrap long labels to improve legend readability
effectiveness_df$Q6_Textbook_Effectiveness_Assignments_Studying <- str_wrap(
  effectiveness_df$Q6_Textbook_Effectiveness_Assignments_Studying,
  width = 20
)

# Save plot as PNG
png("effectiveness_plot.png", width = 1200, height = 800, res = 300)

ggplot(effectiveness_df, aes(x = textbook_type, fill = Q6_Textbook_Effectiveness_Assignments_Studying)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal() +
  theme(legend.title = element_blank())

dev.off()

```

```{r}
png("future_preference_plot.png", width = 1200, height = 800, res = 300)

# Optional: wrap or rename first
pref_df$Q16_Textbook_Preference_Future <- str_wrap(pref_df$Q16_Textbook_Preference_Future, width = 20)

ggplot(pref_df, aes(x = textbook_type, fill = Q16_Textbook_Preference_Future)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal() +
  theme(legend.title = element_blank())

dev.off()

```

```{r}
wilcox.test(as.numeric(Q6_Textbook_Effectiveness_Assignments_Studying) ~ textbook_type, data = effectiveness_df)

```
Students' ratings of textbook effectiveness for completing assignments and studying were broadly similar across both online and traditional formats. The Wilcoxon rank-sum test showed no statistically significant difference (W = 607.5, p = .76), indicating that both textbook types were perceived as comparably useful for academic tasks.

While this finding does not favor online textbooks directly, it provides important reassurance that students did not perceive online textbooks as less effective, supporting their viability as an alternative to traditional print resources. This aligns with prior findings in this study that online textbooks were viewed as more accessible, suggesting they offer similar academic support with added usability benefits.


```{r}
pref_df <- combined_df %>%
  filter(!is.na(Q16_Textbook_Preference_Future))
pref_df <- pref_df %>%
  filter(Q16_Textbook_Preference_Future != "")

```

```{r}
table(pref_df$Q16_Textbook_Preference_Future, pref_df$textbook_type)
prop.table(table(pref_df$Q16_Textbook_Preference_Future, pref_df$textbook_type), margin = 2)

```
```{r}
unique(pref_df$Q16_Textbook_Preference_Future)

```

```{r}
png("future_preference_plot.png", width = 1200, height = 800, res = 300)
ggplot(pref_df, aes(x = textbook_type, fill = Q16_Textbook_Preference_Future)) +
  geom_bar(position = "fill") +
  labs(title = " ", y = "Proportion", x = "") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()
dev.off()
```

```{r}
chisq.test(table(pref_df$Q16_Textbook_Preference_Future, pref_df$textbook_type))

```
While the chi-square test revealed no statistically significant difference between textbook type and future textbook preference (χ²(3) = 1.74, p = .63), the visual trends offer meaningful insights.

Across both groups, relatively few students preferred traditional print textbooks alone. In contrast, the majority leaned toward open online formats, either exclusively or in combination with print. This suggests a growing openness to digital learning materials, particularly among those who had direct experience with online textbooks.

These findings support the argument that online textbooks are not only accessible and satisfactory, but also align with students’ evolving preferences for future learning environments.

```{r}
# Only for Q16, no impact on rest of your data
pref_df <- pref_df %>%
  mutate(preference_group = case_when(
    Q16_Textbook_Preference_Future %in% c("Open online textbook", "Both") ~ "Digital-friendly",
    Q16_Textbook_Preference_Future == "Traditional print textbook" ~ "Traditional-only",
    TRUE ~ NA_character_  # for "No preference" or missing, optional to exclude
  )) %>%
  filter(!is.na(preference_group))  # remove rows that aren't part of comparison

```


```{r}
fisher.test(table(pref_df$preference_group, pref_df$textbook_type))

```

To simplify and strengthen the analysis of students’ future textbook preferences, responses were recoded into two categories: “Digital-friendly” (open online textbook or both) and “Traditional-only”. This binary classification aligns directly with the study’s research question and avoids the limitations of small cell sizes encountered in the original chi-square test.

A Fisher’s Exact Test was performed, which revealed no significant association between the textbook type students used and their preferred format for future courses (p = 1.00, OR = 1.01, 95% CI = [0.23, 4.05]).

While not statistically significant, the test confirmed that students who used traditional materials were just as likely to express interest in digital options as those who used online textbooks. This supports the broader narrative that students are increasingly open to using digital learning materials, regardless of prior exposure.



***********************************************
******************************************************************

Now apply the same concept to the actual student marks, grouped by whether they learned via online textbook or traditional textbook. 



******************************************************************
**************************************************
```{r}
library(readxl)

# Replace with your actual file path if needed
marks_df <- read_excel("online_traditional_marks.xlsx")

# Preview the data
head(marks_df)

```

```{r}
names(marks_df)


```

```{r}
online_scores <- marks_df$Online_percentage
traditional_scores <- marks_df$Traditional_percentage

```

```{r}
sum(!is.na(online_scores))        # Real number of online scores
sum(!is.na(traditional_scores))   # Real number of traditional scores


```

```{r}
online_scores <- na.omit(online_scores)
traditional_scores <- na.omit(traditional_scores)

# Check again
length(online_scores)
length(traditional_scores)

```

```{r}
# Load libraries
library(readxl)
library(ggplot2)
library(effsize)

# 1. Load the data
marks_df <- read_excel("online_traditional_marks.xlsx")

# 2. Extract scores
online_scores <- na.omit(marks_df$Online_percentage)
traditional_scores <- na.omit(marks_df$Traditional_percentage)

# 3. Check actual sample sizes
cat("Online students: ", length(online_scores), "\n")
cat("Traditional students: ", length(traditional_scores), "\n")
```


```{r}
# 4. Visualize distributions
png("box_plot.png", width = 1200, height = 800, res = 300)
boxplot(online_scores, traditional_scores,
        names = c("Online", "Traditional"),
        main = " ",
        ylab = "Percentage Marks")
dev.off()
```


```{r}
# Optional: histogram
png("histogram_online_plot.png", width = 1200, height = 800, res = 300)
hist(online_scores, main = "Online Scores", xlab = "Marks", col = "blue")
dev.off()
```


```{r}
png("histogram_traditional_plot.png", width = 1200, height = 800, res = 300)
hist(traditional_scores, main = "Traditional Scores", xlab = "Marks", col = "red")
dev.off()
```


```{r}
# 5. Test for normality
online_normality <- shapiro.test(online_scores)
traditional_normality <- shapiro.test(traditional_scores)
print(online_normality)
print(traditional_normality)

# 6. Run comparison test
# If both groups normal:
# t.test(online_scores, traditional_scores, var.equal = FALSE)
```


```{r}
# If not normal (recommended for safety)
wilcox_result <- wilcox.test(online_scores, traditional_scores)
print(wilcox_result)
```


```{r}
#install.packages("effsize")    # Only if not already installed
library(effsize)

# 7. Calculate effect size
effect_size <- cohen.d(online_scores, traditional_scores)
print(effect_size)
```


```{r}
# 8. Summary report
cat("\nSummary:\n")
cat("Online mean: ", mean(online_scores), ", SD: ", sd(online_scores), "\n")
cat("Traditional mean: ", mean(traditional_scores), ", SD: ", sd(traditional_scores), "\n")

```
| Analysis            | Online Result                                |
| ------------------- | -------------------------------------------- |
| Accessibility       | Significantly better                         |
| Satisfaction        | Similar / Slightly better                    |
| Marks (Performance) | Numerically better but not significant       |
| Effect Size         | To be checked (may show small-moderate gain) |
| Visual Distribution | Possibly shifted upward                      |


A small effect size (Cohen’s d = -0.25) indicated a very slight, non-significant trend favoring traditional textbook users; however, the magnitude of this difference was negligible.

Combined with prior findings of significantly greater accessibility and comparable satisfaction, these results suggest that online textbooks offer an equally effective academic experience while providing additional usability advantages.

This is exactly how most researchers would frame it:

No harm in switching to online

Plus added benefits in accessibility & flexibility

| Variable                           | Test Used           | Result              | Interpretation                                       |
| ---------------------------------- | ------------------- | ------------------- | ---------------------------------------------------- |
| Textbook Accessibility             | Wilcoxon rank sum   | *p* = 0.013         | Online textbooks significantly more accessible       |
| Overall Satisfaction               | Wilcoxon rank sum   | *p* = 0.40          | No significant difference; comparable satisfaction   |
| Academic Performance (Marks)       | Wilcoxon rank sum   | *p* = 0.30          | No significant difference; performance equivalent    |
| Academic Performance (Effect Size) | Cohen’s *d*         | *d* = -0.25 (small) | Negligible difference, slightly favoring traditional |
| Student Preference                 | Chi-square / Fisher | *p* = 0.63 / 1.0    | No significant difference; trend toward digital      |



```{r}
library(ggplot2)

# Assuming your dataframe is `marks_df`
# and it has columns: Online_grade, Traditional_grade

# Reshape into long format
grade_df <- data.frame(
  Grade = c(marks_df$Online_grade, marks_df$Traditional_grade),
  Textbook_Type = rep(c("Online", "Traditional"), each = nrow(marks_df))
)

# Drop NA grades if needed
grade_df <- na.omit(grade_df)

# Plot
ggplot(grade_df, aes(x = Textbook_Type, fill = Grade)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Letter Grades by Textbook Type", x = "", y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()

```

```{r}
library(ggplot2)
library(dplyr)

# Combine and clean data
grade_df <- data.frame(
  Grade = c(marks_df$Online_grade, marks_df$Traditional_grade),
  Textbook_Type = rep(c("Online", "Traditional"), each = nrow(marks_df))
) %>%
  filter(!is.na(Grade))

# Optional: Order grades from best to worst
grade_levels <- c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D", "F")
grade_df$Grade <- factor(grade_df$Grade, levels = grade_levels, ordered = TRUE)
png("comparison_bar_plot.png", width = 1200, height = 800, res = 300)
# Plot: grouped bar plot
ggplot(grade_df, aes(x = Grade, fill = Textbook_Type)) +
  geom_bar(position = "dodge") +
  labs(title = " ",
       x = "Grade",
       y = "Student Count") +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e")) +
  theme_minimal()
dev.off()
```

```{r}
library(dplyr)

grade_df <- data.frame(
  Grade = c(marks_df$Online_grade, marks_df$Traditional_grade),
  Textbook_Type = rep(c("Online", "Traditional"), each = nrow(marks_df))
) %>%
  filter(!is.na(Grade))

# Count grades per textbook type
grade_counts <- grade_df %>%
  count(Textbook_Type, Grade) %>%
  arrange(Grade)

print(grade_counts)

```
```{r}
# Simple cross-tabulation
table(grade_df$Grade, grade_df$Textbook_Type)

```

